---
title: "Statistical Models in R: Day 1"
subtitle: "Analysis of Variance (ANOVA) in R"
author: |
  | Tiffany Timbers
  | Applied Statistics and Data Science Group
  | UBC Statistics
date: February 6, 2017
fontsize: 10pt
output:
  beamer_presentation:
    pandoc_args: [
      # font setup
      '--latex-engine=xelatex',
      '--variable=header-includes:
        \usepackage{fontspec}
        \defaultfontfeatures{Mapping=tex-text}
        %\setsansfont[Ligatures={Common}]{Calibri}',
      # slide footer
      '--variable=header-includes:
        \setbeamercolor{footlinecolor}{fg=black,bg=lightgray}
        \setbeamertemplate{footline}{
          \ifnum\value{framenumber}>1
          \begin{beamercolorbox}[ht=2.5ex,dp=1ex,wd=\paperwidth]{footlinecolor}
          \makebox[.2\paperwidth][l]{
            \quad Introduction to R
          }\makebox[.6\paperwidth][c]{
            Applied Statistics and Data Science Group
          }\makebox[.2\paperwidth][r]{
            Day 1\quad
          }\end{beamercolorbox}
          \else\vspace{10pt}\fi
        }
        \setbeamertemplate{navigation symbols}{}
        \setbeamertemplate{itemize items}[circle]
        \hypersetup{colorlinks,linkcolor=,urlcolor=magenta}',
      # workaround for multi-column slides
      '--variable=header-includes:\newcommand{\columnsbegin}{\begin{columns}}',
      '--variable=header-includes:\newcommand{\columnsend}{\end{columns}}'
    ]
    fonttheme: professionalfonts
---

```{r load libraries, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
```

# ANOVA - when to use it?

## Types of data

- response/dependent variable ($Y$) is quantitative
  - examples of quantitative variables:
    - height
    - salary
    - number of offspring

- your explanatory/independent variable(s) ($X$'s) are categorical
  - examples of categorical variables:
    - eye color
    - sex
    - genotype at a given locus

## Examples of Cases where ANOVA would be used

1. Does diet has an effect on weight gain?
    - response variable = weight gain (*e.g.*, kg)
    - explanatory variable = type of diet (*e.g.*, low vs. medium vs. high sugar)
2. Does the type sexual relationship practiced influence the fitness of male Red-winged Blackbirds?
    - response variable = fitness of male bird (*e.g.*, # eggs laid)
    - explanatory variable =  sexual relationship (*e.g.*, monagamy vs. polygamy)

# ANOVA - how it works?

## Example case:

- Does diet has an effect on chick weight after 6 weeks?
- response variable == chick weight (grams)
- explanatory variable == type of diet (4 levels)

$$H_{0} : \mu_{caesin} = \mu_{horsebean} = \mu_{linseed} = \mu_{meatmeal} = \mu_{soybean} = \mu_{sunflower}$$

$H_{A}$ : at least one group's population mean differs from that of the other groups

```{r chick weight plot, echo = FALSE, fig.height = 3}
set.seed(1234)
ggplot(chickwts, aes(x = feed, y = weight, color = feed)) +
  geom_jitter(position = position_jitter(0.15),
              alpha = 0.7) +
  stat_summary(fun.y = mean,
               geom = "point",
               shape = 5,
               size = 5,
               aes(group = feed)) +
  theme(legend.position="none") +
  xlab("Type of diet") +
  ylab("Weight gain over 6 weeks (grams)") +
  ylim(0, 450)
```

## The gist of the math behind it

```{r chick weight plot repeat, echo = FALSE, fig.height = 3}
set.seed(1234)
ggplot(chickwts, aes(x = feed, y = weight, color = feed)) +
  geom_jitter(position = position_jitter(0.15),
              alpha = 0.7) +
  stat_summary(fun.y = mean,
               geom = "point",
               shape = 5,
               size = 5,
               aes(group = feed)) +
  theme(legend.position="none") +
  xlab("Type of diet") +
  ylab("Weight gain over 6 weeks (grams)") +
  ylim(0, 450)
```

To calculate a test statistic (F-statistic), we compare the **between** group variation with the **within** group variation

$$F = MSB/MSW$$

where, $MSB$ = Mean Square Between & $MSW$ = Mean Square Within
\hfill \break
Essentially, if there is greater variation between the groups than within the groups we will get a large test statistic value (and correspondingly, a small p-value) and reject that null hypothesis ($H_0$: population means of all groups are equal).

## Want to know more?

Watch this excellent series of videos from Khan Academy where they perform ANOVA by hand for a simple case:

https://www.khanacademy.org/math/statistics-probability/analysis-of-variance-anova-library

## Key assumptions / rules of thumb

- ANOVA is robust to the non-normality of sample data

- Balanced ANOVA (equal sample size between groups) is robust to uneuqal variance

- ANOVA is sensitive to sample indpendence

# Syntax for ANOVA in R

## Laying out your dataset

- Most statistical functions in R work best with a "tidy" dataset
    - tidy data def'n: *"each variable is a column, each observation is a row"*

- For our case (chick feed example), this means 2 columns, one for the response variable and one for the explanatory variable:

```{r example of tidy data, echo = FALSE}
set.seed(1234)
kable(sample_n(chickwts, 6), row.names = FALSE)
```

for more information on tidy data see: http://vita.had.co.nz/papers/tidy-data.pdf

## The `aov` function

- To perform ANOVA in R we can use the `aov` function
- `aov` requires the following arguments:
    - `formula`
    - `data`

Here is an example call to `aov` using our chick feed case example:

```{r example call to aov}
chick_feed_model <- aov(weight ~ feed, data = chickwts)
```

To get the results from the ANOVA analysis performed by `aov` in a nice tidy data frame, I recomend using the `tidy` function from R's `broom` package:

```{r example call to tidy on aov object}
library(broom)
broom::tidy(chick_feed_model)
```

## `summary` - another way to get results from `aov`

Traditionally the `summary` function was used. This provides a nice print output, but unlike tidy, the results are not contained in a data frame, and thus more tricky to access:

```{r example call to summary on aov object}
summary(chick_feed_model)
```
